\chapter{Conclusion and Future Work}
\label{ch:concl}
The work reported in this thesis was conceived to bring contributions in three aspects. First, in order to mine data from social networks (e-mail lists and others), a number of tools and procedures to deal with large amounts of data were developed. More specific details will be mentioned in connection with linked data and visualization tools. The second contribution is associated with a thorough analysis of the dynamics of social networks represented by email lists. The third contribution arose from further exploring the findings in the second contribution. Having found that a partition of networks into hubs, intermediary and peripheral participants was very stable over time, we then discovered that the pattern of language usage could be typical of the classes of participants. In the following, the overall conclusions from all of these contributions are described in further detail. 

\section{Stability of topology in the networks}

The very small standard deviations of principal components formation
(see Sections~\ref{sec:pca} and~\ref{prevalence}),
the presence of the Erd\"os sectors even in networks with
few participants (see Sections~\ref{sectioning} and~\ref{subsec:pih}),
and the recurrent activity patterns along different timescales (see Sections~\ref{sec:mtime} and~\ref{constDisc}),
go a step further in characterizing scale-free networks in the context
of the interaction of human individuals.
Furthermore, the importance of symmetry-related metrics,
with respect to dispersion of the system in the topological measures space,
might add to the current understanding of key-differences between digraphs and
undirected graphs in complex networks.
Noteworthy is also the very stable fraction of participants in each Erd\"os sector when the network reaches more than 200 participants.
Benchmarks were derived from email list networks
and the supplied analysis of
networks from Facebook,
Twitter and ParticipaBR in Appendix~\ref{ap:textd} might allow for hypothesizing
about the generality of these characteristics.

\section{Textual analysis final remarks}\label{sec:remarks}
%Human interaction networks yield diverse linguistic peculiarities reported by its members.
This is a first systematic exploration of the relation between topological and textual
metrics in human interaction networks, as far the author knows.
Different textual features were scrutinized and were found to present
salient patterns, especially in relation to topological measures and the Erd\"os sectors.
Furthermore, results show that peripheral participants use more nouns while hubs use more verbs,
which suggests that less connected participants bring content and concepts,
while hubs propose action on them.
Such findings have potential applications in the collection and diffusion of information,
resources recommendation in linked data contexts,
and open processes of document elaboration and refinement~\cite{ensaio,ops,opa,stab,pnud4,pnud3}.

Most importantly, it is reasonable to conclude, from all the distinct textual characteristics
between the Erd\"os sectors, that, as a rule of thumb, the texts from each of the sectors differ.
Surely there should be exceptions and it is a fact that we left out of the analysis
more subtle textual aspects e.g. those related to low percentages (<5\%) or to small
differences.
These might be the subject of future studies.

\section{Linked data final remarks}
The database presented in this thesis
constitutes a large database with diverse provenance.
Even so, the database should be expanded upon need or requests from feedback.
All data should be available online in the \url{http://linkedopensocialdata.org}
address in the near future to fulfill the purpose of being a common
repertoire in current research.
One should reach the diagrams and tables of the 
articles produced in this research~\cite{stab,rcText,versinus,losd}
for further directions
on the available structures and for an overview complement.
Complementary linked data/semantic web ontologies, triplification routines
and documentation are in Appendix~\ref{ap:more}.

\section{Network visualization final remarks}
Versinus was useful both to understanding the interaction networks and to
grasp important dynamic graph visualization features.
The mapping to musical audio of network characteristics added to
the allure of Versinus but did not ease the consideration of the structures
(at least not in a conscious way).
The visualization was relevant for presenting the research in a number of occasions,
specially to exhibit the apparent disorder of the networks,
and that the agents participates with a constantly changing concentration of activity.
These facts expose how counter-intuitive are the stability we report in this thesis.
Less important for our results, but not for the course of this research,
are the static and dynamic network visualizations presented in Appendix~\ref{ap:more}.

\section{Future work}\label{sec:fw}

Further work should expand the analysis to include
more types of networks and more metrics.
The data and software needed to attain these results
received dedicated and in-depth
documentation as they enable a greater level of transparency
and work share,
which is adequate for both benchmarking
and specifically for the study of systems constituted
by human individuals (see Section~\ref{sec:data}).
The derived typology of hub, intermediary and peripheral participants
has been applied for semantic web and participatory democracy efforts,
and these developments might be enhanced to yield scientific knowledge~\cite{pnud4,opa,losd}.
Also, we plan to further explore and publish the audiovisualizations
used for this research~\cite{versinus,animacoes} and
the linguistic differences found in each of the Erd\"os sectors~\cite{rcText}.

Similarity measures of texts in message-response threads have been conceived, 
and some results should be organized in the near future.
In this respect, there are two core hypotheses obtained from recent experiments:
\begin{itemize}
\item the existence of information ``ducts'', observable through similarity measures.
These might coincide with asymmetries of edges between vertex pairs,
with homophily or with message-response threads, to point just a few possibilities.
\item Valuable insights might be obtained from the self-similarity of messages by the same author,
of messages sent at the same period of the day, etc.
This includes incidences of word sizes, incidences of tags and morphosyntactic classes,
incidences of particular Wordnet synset characteristics and distances.
\end{itemize}

The current results suggest that diversity and self-similarity should vary with respect to connectivity. In the literature, it is usually assumed 
that the periphery holds greater diversity~\cite{easly},
which can be further verified, for example through the diversity of entries (e.g. tokens, sentence sizes).

Other potential next steps are:
\begin{itemize}
\item The observation of most incident words and word types,
such as words related to cursing, food or body parts.
\item Interpretation of the constant difference found from incident and existent tokens histograms,
mentioned in Section~\ref{subsec:sii}.
\item Extend word class observations, e.g. to include plurals, gender, common prefixes and suffixes.
\item The observation of date and time in relation to textual production of interaction networks and
to activity characteristics (e.g. dispersion of sent time along the day or weekdays).
\item A careful analysis of each textual feature distribution which is likely to reveal multimodal outlines and other non-trivial characteristics.
\item Extend the analysis of textual measurements to the windowed approach along the timelines, where hub, peripheral and intermediary sectors were topologically characterized~\cite{stab}.
\item For ELE list, the more connected the sector, the longer the messages are.
This is the inverse of what was found in the other lists,
and was considered a peculiarity of the culture bonded with the political subject of ELE list.
This hypothesis should be further verified.
\item Tackle the same analysis on networks with languages other than English.
This is especially important for applications~\cite{ensaio}
and should rely on dedicated implementation of 
tokenization, lemmatization and attribution of POS tags.
\item Observe a broader set of human interaction networks and the resulting types
of networks and participants with respect to topological and textual features.
\item Analyze interaction networks from other platforms such as LinkedIn, Diáspora, etc.
\item Sentiment analysis was not used in this work but might be pertinent since the topic has received considerable attention from the scientific literature but has not included topological features as far as we know.
\item This thesis focused on differences of texts among Erd\"os sectors but we envisage that comparison of texts from social networks against canonical texts (e.g. Shakespeare or the King James Bible) might yield other powerful insights.
\item The significant differences found from the texts of the sectors raises the question as to why we are not conscious of these differences.
One possibility is that they are part of our instinctive and unconscious social coordination and this hypothesis might be the goal of future efforts.
\item Although we used the fact that social networks exhibit a scale-free outline,
	we were not interested to know if the networks analyzed they were scale-free.
		This might be an interesting endeavor given the diversity of the data we gathered.
\item Consider communities in the analysis: how do they evolve? Does e.g. the inclusion of a modularity index add to the acquired knowledge?
\item To analyze the dynamics of individual participants: how they transition between one (Erdös) sectors, when are their activity stable, etc.
\item Topic detection can be very interesting to tackle from the analysis proposed in this thesis:
	e.g. in one thread, are topics preserved? Do Intermediary have more recurrent topics than hubs?
\item Our analysis considered mainly local measures (and global statistic derived from them), with the exception of betweenness centrality.
	New insights might be obtained by the inclusion of more non-local measures.
\item A thorough consideration of the analysis made in this thesis within the literature of dynamic networks.
\item The inclusion of more sophisticated analyses of text, such as those used in the Coh-Metrix framework for text complexity~\cite{coh}.
\item We performed measurements derived from TF-IDF (term frequency-inverse document frequency),
	applied to the analysis of texts from the Erd\"os sectors.
\item Using machine learning to quantify the separability of the Erdös sectors with respect to their respective texts.
		We envisage their careful inclusion in further studies.
	\item This research included many exploratory endeavors which are included in Appendix~\ref{ap:more}.
		They were important in enhancing our understanding of human social structures,
		have documentations and software, and might benefit from more scientific presentations.

\item We performed multiscale analyzes by going through networks derived by tenths to thousands of messages.
	This enabled us to better evidence stability, and the presence of Erd\"os sectors in small networks, 
		but did not yield other significant knowledge.
		We hypothesize that it will meaningful to explore the upper boundary for which our results hold,
		i.e. to observe networks derived from a large enough number of messages for the structures to changes.
		It might also be interesting to measure how well the networks of different sizes fit the power-law distribution
		of connectivity.
\end{itemize}

% Multiscale analysis
% Inclusion of more metrics


% Versinus


% Collection and diffusion of information


